---
title: "Clean, Analyze, Visualize"
output:
  html_document:
    df_print: paged
---

Let's work through a sports example since it's March Madness and my bracket is already busted. 

**Background:** I went to Marquette Univeristy and my freshman year was Jimmy Butler's final season (2010-11 was awesome!)

So this is the data we'll work with: 

**Marquette team data**

+ College basketball gamelogs for [Marquette University's 2010-11 season](https://www.sports-reference.com/cbb/schools/marquette/2011-gamelogs.html).

**And the gamelogs for each individual player:** 

+ [Darius Johson-Odom](https://www.sports-reference.com/cbb/players/darius-johnson-odom-1/gamelog/2011)

+ [Jimmy Butler](https://www.sports-reference.com/cbb/players/jimmy-butler-1/gamelog/2011)

+ [Jae Crowder](https://www.sports-reference.com/cbb/players/jae-crowder-1/gamelog/2011)

+ [Dwight Buycks](https://www.sports-reference.com/cbb/players/dwight-buycks-1/gamelog/2011)

+ [Chris Otule](https://www.sports-reference.com/cbb/players/chris-otule-1/gamelog/2011)

+ [Vander Blue](https://www.sports-reference.com/cbb/players/vander-blue-1/gamelog/2011)

+ [Davante Gardner](https://www.sports-reference.com/cbb/players/davante-gardner-1/gamelog/2011)

+ [Junior Cadougan](https://www.sports-reference.com/cbb/players/junior-cadougan-1/gamelog/2011)

+ [Joe Fulce](https://www.sports-reference.com/cbb/players/joe-fulce-1/gamelog/2011)

+ [Erik Williams](https://www.sports-reference.com/cbb/players/erik-williams-1/gamelog/2011)

+ [Jamail Jones](https://www.sports-reference.com/cbb/players/jamail-jones-1/gamelog/2011)

+ [Reggie Smith](https://www.sports-reference.com/cbb/players/reggie-smith-1/gamelog/2011)

+ [Robert Frozena](https://www.sports-reference.com/cbb/players/robert-frozena-1/gamelog/2011)

**Some of the later visualizations in this example will work through this template from Matt Waite at NICAR 2019:** [https://mattwaite.github.io/nicar19/CourseMaterials.html](https://mattwaite.github.io/nicar19/CourseMaterials.html)

+ [All team game logs for the 2018-19 season](https://github.com/mattwaite/NICAR19SportsDataViz)

_**All datasets sourced from Sports Reference.**_

-----

## Goal

Clean, analyze and visualize these datasets. To do that we'll work with R's `tidyverse` library.

To save on typing or copying and pasting all lines of code in this feel free to open the `final-example.Rmd` [from my github repo](https://github.com/erincaughey/usat-nicar-presentation) file in RStudio and run the code chunks.

```{r libraries, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
#uncomment and run install code if tidyverse is not already running
#install.packages("tidyverse")
library(tidyverse)

# optional library could be useful for plotting later
#install.packages("lubridate") #helps us reformat the date column
library(lubridate)

```

We'll work toward completeing 3 different graphcis. I'll denote which cleaning, analyzing and plotting steps go with each dataset using these labels: 

1. <span style="background-color:#506987; padding: 3px; border-radius: 3px; color: #fff;">MU team</span>

2. <span style="background-color:#ff6347; padding: 3px; border-radius: 3px; color: #fff;">MU individuals</span>

3. <span style="background-color:#89A46F; padding: 3px; border-radius: 3px; color: #fff;">All teams</span>

-----

## Clean

<span style="background-color:#89A46F; padding: 3px; border-radius: 3px; color: #fff;">All teams</span>

We're going to load this and forget about it for a minute because Matt Waite did the dirty work of cleaning this for us already. 

```{r all team logs, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
all_teams <- read_csv("data/1819-all-logs.csv")
```

<span style="background-color:#506987; padding: 3px; border-radius: 3px; color: #fff;">MU team</span>

These are gamelogs for [Marquette University's 2010-11 season](https://www.sports-reference.com/cbb/schools/marquette/2011-gamelogs.html). There's an option to download the file under **"Share &more"**. Select **"Get Table as CSV"**, copy the text on the page, paste into a text editor and then save the file as `mu-10-11-gamelong.csv`. _Or find the necessary file in the github repo data folder._

```{r mu gamelogs clean, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
logs <- read_csv("data/mu-10-11-gamelog.csv", skip=1)
# we use skip = 1 becuase there is an extra row of label data that we don't need.
# Let's look at the column names or variables for this data
colnames(logs)

# create a new dataframe 'cleaned_logs' here and rename each column, separate by comma
# rename(dataframe, new_name = old_name, new_name = old_name, new_name = old_name...) 
clean_logs <- 
  rename(logs, game = G, date = Date, home_away = X3, opponent = Opp, w_l = `W/L`, team_score = Tm, opponent_score = Opp_1, team_FG = FG, team_FGA = FGA, team_FGpct = `FG%`, team_3P = `3P`, team_3PA = `3PA`, team_3Ppct = `3P%`, team_FT = FT, team_FTpct = `FT%`, team_ORB = ORB, team_TRB = TRB, team_AST = AST, team_STL = STL, team_BLK = BLK, team_TOV = TOV, team_PF = PF, opponent_FG = FG_1, opponent_FGA = FGA_1, opponent_FGpct = `FG%_1`, opponent_3P = `3P_1`, opponent_3PA = `3PA_1`, opponent_3Ppct = `3P%_1`, opponent_FT = FT_1, opponent_FTpct = `FT%_1`, opponent_ORB = ORB_1, opponent_TRB = TRB_1, opponent_AST = AST_1, opponent_STL = STL_1, opponent_BLK = BLK_1, opponent_TOV = TOV_1, opponent_PF = PF_1)

# remove the blank separater column and reassign to clean_logs
clean_logs <- clean_logs %>% select(-X24)
```

<span style="background-color:#ff6347; padding: 3px; border-radius: 3px; color: #fff;">MU individuals</span>

Now let's combine and clean each of the players gamelog data from Marquette University's 2010-11 roster. 

I followed the same steps as above to save a `.csv` file for each individual player on the team. This is the hard way but also a good example of how to combine these files into one since they are all formatted the same way.

```{r players clean and join, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
# run each group of functions separately

odom <- read_csv("mu-players/djo-1011.csv") 
butler <- read_csv("mu-players/jbutler-1011.csv")
crowder <- read_csv("mu-players/crowder-1011.csv")
buycks <- read_csv("mu-players/buycks-1011.csv")
otule <- read_csv("mu-players/otule-1011.csv")
blue <- read_csv("mu-players/blue-1011.csv")
gardner <- read_csv("mu-players/gardner-1011.csv")
cadougan <- read_csv("mu-players/cadougan-1011.csv")
fulce <- read_csv("mu-players/fulce-1011.csv")
williams <- read_csv("mu-players/williams-1011.csv")
jones <- read_csv("mu-players/jones-1011.csv")
smith <- read_csv("mu-players/smith-1011.csv")
frozena <- read_csv("mu-players/frozena-1011.csv")

# add a column for the players name to join easier
odom$player <- "Darius Johnson-Odom"
butler$player <- "Jimmy Butler"
crowder$player <- "Jae Crowder"
buycks$player <- "Dwight Buycks"
otule$player <- "Chris Otule"
blue$player <- "Vander Blue"
gardner$player <- "Davante Gardner"
cadougan$player <- "Junior Cadougan"
fulce$player <- "Joe Fulce"
williams$player <- "Erik Williams"
jones$player <- "Jamail Jones"
smith$player <- "Reggie Smith"
frozena$player <- "Robert Frozena"

#TIP: can re-export a csv updated with name column to replace current one if needed
#write_csv(odom, "mu-players/djo-1011.csv") 

mu_players <- merge(odom, butler, all=TRUE) %>% 
  merge(., crowder, all = TRUE) %>% 
  merge(., buycks, all = TRUE) %>%
  merge(., otule, all = TRUE) %>% 
  merge(., blue, all = TRUE) %>% 
  merge(., gardner, all = TRUE) %>% 
  merge(., cadougan, all = TRUE) %>% 
  merge(., fulce, all = TRUE) %>% 
  merge(., williams, all = TRUE) %>% 
  merge(., jones, all = TRUE) %>% 
  merge(., smith, all = TRUE) %>% 
  merge(., frozena, all = TRUE)

# rename some columns and filter out each player's overall season row at the bottom
# but before we remove overall season totals let's bring those into a separate table for some analysis
player_totals <- mu_players %>% filter(is.na(Rk)) %>% rename(total_games = Date)
player_totals <- select(player_totals, -c(1, 3:7))

mu_players <- mu_players %>% rename(game = Rk, date = Date, opponent = Opponent, type = Type, home_away = X4, win_loss = X7) %>% filter(., !is.na(game))

# Let's export the mu_players csv to save our cleaning.
write_csv(mu_players, "mu-players/all.csv")

```

-----

## Analyze

<span style="background-color:#506987; padding: 3px; border-radius: 3px; color: #fff;">MU team</span>

First, let's go back to our Marquette team season logs and look at the score differential.

**TIP:** The `%>%` is the pipe symbol. It help you apply multiple commands to the same dataframe by stringing them together.

Here we do some math by adding a new column with `mutate()` called "differential" to show how much the team wins or loses by. And then sort that column in decending order. `arrange()` is `dplyr` version of `sort()`.

```{r gamelog differentials, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
clean_logs %>% 
  mutate(differential = team_score-opponent_score) %>% 
  arrange(desc(differential)) -> arranged_logs
```

We'll come back to that later when we plot that dataframe.

<span style="background-color:#ff6347; padding: 3px; border-radius: 3px; color: #fff;">MU individuals</span>

Next, let's jump to individual gamelogs and compare player's overall totals throughout the season.

This is why I used `filter()` to grab the final row of each player's gamelogs. It includes the cummulative stats for the amount of games each Marquette team member played throughout the 2010-11 season.

An obvious assumption is that the more games you play the higher totals and season averages a player will have. We can use the `arrange()` function to check that correlation. 

```{r player analysis on totals, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
arrange(player_totals, desc(PTS))
# now replace "PTS" above with any of the other column names like AST, BLK, 3P, etc. 
# remember if you try and arrange by any of the columns that start with a number you need to use backticks (ex: `3P`) so R will understand.
```

Becasue we kept each individual player's data we can run some comparisons between teammates throughout the season too. 

For example let's just use `arrange` again see who scored the most points per game throughout the season. You'll have to scroll all the way to the right to see the PTS and player variables.

```{r arrange, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
arrange(mu_players, desc(PTS))
# this shows Jimmy Butler, Darius Johnson-Odom and Jae Crowder had some of the highest points per game
```

We can also use filter to see how many players scored above double digits in a game. 
Remember we have 383 rows of data. 

```{r arrange and filter, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
more_than_ten <- filter(mu_players, PTS >= 10) %>% arrange(., desc(PTS))
```

This table shows 133 rows or **records** of instances where Marquette players scored greater than or equal to 10 points during a game

If you scan through you can see Darius Johnson-Odom scored over 10 points per game A LOT. 

Let's go one step further and find the average amount of games DJO scored over 10 points in a game. 

```{r Ex. DJO PPG, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
filter(more_than_ten, player == "Darius Johnson-Odom")
```
That's 33 records. Or 33 games. And from our `player_totals` dataframe we know that he played a total of 37 games. Simple average: `33 / 37 = .8918919` or of the 37 games Darius Johnson-Odom played he scored 10 points or more in 89% of those games.

But of course there's a simpler way to find simple statistics on individual players or datasets. The `summary()` function.

With this we would be able to check a players mean for various statistics to see how they might compare to other players.

```{r summary, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
summary(odom)
# ^ replace "odom" with other player tables to compare
```

-----

## Visualizing

<span style="background-color:#506987; padding: 3px; border-radius: 3px; color: #fff;">MU team</span>

First let's again revisit the Marquette team gamelogs and create a bar graph to show wins and losses for Marquette's 2010-11 season based on the point differential.

```{r bar chart, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
ggplot() + geom_bar(data=arranged_logs, aes(date, weight=differential))
```

<span style="background-color:#ff6347; padding: 3px; border-radius: 3px; color: #fff;">MU individuals</span>

And now a look at the individual Marquette players for that season.

Jae Crowder was my favorite player in 2010. So let's start off by plotting how many points he scored in each game throughout the season with `ggplot` in `tidyverse`

First, I need to remove the total points from Jae's individual gamelog table since I only removed them in the combined `mu_players` dataframe. 

You'll see labeling and color changes specified in this one. I'll dig into this a little later.

```{r Crowder plot, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
# ymd() function uses "lubridate" package so load that library if needed.

jae_clean <- filter(crowder, !is.na(Rk)) %>% mutate(date = ymd(Date))

# now let's plot and add labels
jae_pts <- ggplot(jae_clean, aes(date, PTS)) +
  geom_point(aes(color = factor(Type))) + 
  labs(title = "Jae Crowder's points per game (2010-11)", y = "Points", x = "Game", caption = "Source: Sports-Reference.com") +
  theme(axis.text.x = element_text(angle=45)) +
  # And let's fix the colors and legend
  scale_color_brewer(name = "Game Type", type = "qual", palette = "Dark2",
                               labels = c("Conference Tourny",
                                          "NCAA Tourny",
                                          "Regular season")) +
  theme_minimal()

jae_pts 

```

How about all the players next to each other?

```{r multiples, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
# now let's plot and add labels
mu_players <- mutate(mu_players, date = ymd(date))
all_players <- ggplot(mu_players, aes(date, PTS)) +
  geom_point(aes(color = factor(type)), size = .8) + 
  labs(title = "Points per game (2010-11)", y = "Points per game", x = "Games played", caption = "Source: Sports-Reference.com") +
  facet_wrap(~ player) + 
  theme_minimal()

all_players
```

<span style="background-color:#89A46F; padding: 3px; border-radius: 3px; color: #fff;">All teams</span>

Now let's step back and look at all NCAA teams 2018 season gamelogs as of the last week in February.

_This is where that first dataset from Matt Waite comes into play._ 

Since we've been talking about Marquette let's look at all team game logs and pull out just those in the Big East conference. 

```{r filter league, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
all_teams %>% filter(Conference == "Big East") -> big_east
```

Let's visualize the field goal percentage across all Big East teams during the 2018-19 season.

```{r lines, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
ggplot() + geom_line(data=big_east, aes(x=Date, y=TeamFGPCT, group=Team))
```

That's messy. Clean it up by highlighting specific teams and layering and setting styles. 

We'll create a new working dataframe with `grouped_by()` and create columns for each team's FG, FG attempts and a calculated FG percentage.

```{r big east FG, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
big_east %>% 
  group_by(Team) %>% 
  summarize(
    season_attempts = sum(TeamFGA),
    season_fg = sum(TeamFG),
    season_pct = season_fg/season_attempts
  ) %>% 
  arrange(desc(season_pct))
```

Using this let's compare Marquette to the team with the best season FG percentage, Creighton.

```{r filter teams, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
mu <- big_east %>% filter(Team == "Marquette Golden Eagles")
cb <- big_east %>% filter(Team == "Creighton Bluejays")

```

And replot them with style: 

```{r layer and style, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
ggplot() + 
  geom_line(data=big_east, aes(x=Date, y=TeamFGPCT, group=Team), color="light grey") +
  geom_line(data=mu, aes(x=Date, y=TeamFGPCT, group=Team), color="gold") +
  geom_line(data=cb, aes(x=Date, y=TeamFGPCT, group=Team), color="blue")
```

Now we can see the teams but it's hard to decipher who's doing better just looking at the FG percentage. 

How about getting more point differentials for those teams in the Big East like we did with just Marquette's 2010-11 season gamelogs?

```{r team differentials, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
big_east_tot <- big_east %>% 
  mutate(differential = TeamScore-OpponentScore) %>% 
  group_by(Team) %>% 
  mutate(totaldifferential = cumsum(differential))

# let's highlight Creighton and Marquette again
muc <- big_east_tot %>% filter(Team == "Marquette Golden Eagles")
cbc <- big_east_tot %>% filter(Team == "Creighton Bluejays")
```

Let's track cumulative change over time using a step chart. You'll notice everything is about the same except we'll use `geom_step` instead of `geom_line`

```{r step chart, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
ggplot() + 
  geom_step(data=big_east_tot, aes(x=Date, y=totaldifferential, group=Team), color="light grey") + 
  geom_step(data=muc, aes(x=Date, y=totaldifferential, group=Team), color="gold") +
  geom_step(data=cbc, aes(x=Date, y=totaldifferential, group=Team), color="blue")
```

You'll see that even though Creighton had an overall better FG pecentage in the Big East, Marquette still outpaces them in scoring differentials throughout the season.

### Finalize the grahpic

For better understanding let's get some readable labels (`labs()`), source lines and a little cleaner style. These can be applied to most any chart in `ggplot`.

We'll start by remaking the chart above:

```{r step chart - with labels, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
final_plot <- ggplot() + 
  geom_step(data=big_east_tot, aes(x=Date, y=totaldifferential, group=Team), color="grey") + 
  geom_step(data=muc, aes(x=Date, y=totaldifferential, group=Team), color="gold") +
  geom_step(data=cbc, aes(x=Date, y=totaldifferential, group=Team), color="blue") +
  
  # and we'll add our labelling into the ggplot structure
  labs(x="Date", y="Score differential", title="The January drive", subtitle="Marquette's new year scoring outpaced Big East contenders", caption="Source: Sports-Reference.com and Matt Waite code") +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size=10),
    axis.title.x = element_blank(),
    axis.text = element_text(size = 7),
    axis.ticks = element_blank()
  ) +
  theme_minimal()

# theme_minimal() takes out the gray grid background and simplifies our axes a little more
final_plot 
```

## Export your work

`ggsave()` will export your most recently run chart in R. 

Export filetype: 
```{r export, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
# PDF or PNG 
# to speficiy a chart add a "plot = plot_name" parameter after the file name.
ggsave("exports/players-plot.pdf", plot = all_players, width = 7, height = 10, units = "in", dpi = 320)
ggsave("exports/players-plot.png", plot = all_players, width = 7, height = 10, units = "in", dpi = 320)

ggsave("exports/big-east-plot.pdf", plot = final_plot, width = 16, height = 9, units = "in", dpi = 320)
ggsave("exports/big-east-step.png", plot = final_plot, width = 16, height = 9, units = "in", dpi = 320)

```

